---
title: "Getting Started: Competing Risks Pipeline"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started: Competing Risks Pipeline}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4,
  eval = requireNamespace("cmprsk", quietly = TRUE)
)
```

```{r setup}
library(omicsuite)
```

## Why not just treat competing events as censoring?

Suppose the event of interest is relapse, but some patients die from an
unrelated cause first. Treating that death as ordinary censoring in a
standard KM or Cox analysis implicitly assumes those patients would have
gone on to relapse eventually, had they lived -- an assumption you can
never check and usually don't believe. `fit_competing_risks_pipeline()`
instead estimates the cumulative incidence directly, keeping those
subjects in the risk set on the correct (subdistribution) timescale.

**The event coding is different here.** `fit_km_pipeline()` and
`fit_coxph_pipeline()` want a binary 0/1 event indicator. This function
wants `event_var` to encode *which* event happened: `0` = censored, `1` =
the event of interest, `2` = a competing event, and so on.

## A simulated example

```{r}
set.seed(1)
n <- 300
dat <- data.frame(
  time  = rexp(n, rate = 0.1),
  event = sample(c(0, 1, 2), n, replace = TRUE, prob = c(0.3, 0.5, 0.2)),
  arm   = factor(sample(c("control", "treatment"), n, replace = TRUE))
)

fit <- fit_competing_risks_pipeline(
  dat, "time", "event",
  group_var  = "arm",
  covariates = "arm"
)
fit
```

## Reading the verdict table

```{r}
fit$verdicts
```

`event_type_breakdown` is worth checking first -- how many subjects landed
in each event code, including censoring. `subdistribution_vs_cause_specific`
is a standing reminder, not a data-dependent result: the Fine-Gray model
here answers a different question than a cause-specific Cox model would,
and the two can point in different directions for the same covariate.
`interpretation[gray_test_event_<code>]` is the group-comparison test per
event type; `interpretation[subdistribution_hr_<term>]` is the Fine-Gray
model's subdistribution hazard ratio, phrased the same way as
`fit_coxph_pipeline()`'s hazard ratios but explicitly labeled as a
*subdistribution* HR so the two don't get conflated in a write-up.

## The cumulative incidence plot

```{r, fig.alt = "Cumulative incidence curves by event type and treatment arm"}
plot(fit, which = "cif_plot")
```

Each event type gets its own line style, each group its own color -- so
you can see at a glance whether the event of interest and the competing
event are moving in the directions you'd expect for each arm. The plot's
caption below it carries the same Gray's-test and subdistribution-HR
interpretation as the verdict table, so the figure stands on its own if
you're pasting it directly into a write-up.
