---
title: "Getting Started: Survival Pipeline"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started: Survival Pipeline}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4
)
```

```{r setup}
library(omicsuite)
```

## Why a pipeline instead of just `coxph()`

Fitting a Cox model is one line of code. Trusting it takes several more:
checking the proportional hazards assumption, checking whether a handful of
observations are driving the estimate, and checking whether continuous
covariates actually enter the model linearly on the log-hazard scale.
`fit_coxph_pipeline()` runs all of that in one call and returns a structured
verdict alongside the model.

## A simulated example

```{r}
set.seed(1)
n <- 300
dat <- data.frame(
  time  = rexp(n, rate = 0.05),
  event = rbinom(n, 1, 0.7),
  arm   = factor(sample(c("control", "treatment"), n, replace = TRUE)),
  age   = rnorm(n, 55, 10)
)

fit <- fit_coxph_pipeline(
  data       = dat,
  time_var   = "time",
  event_var  = "event",
  covariates = "arm",
  adjust_for = "age"
)

fit
```

## Reading the verdict table

```{r}
fit$verdicts
```

Each row is one diagnostic check. `"pass"` means no evidence of a problem;
`"flagged"` means the check crossed the threshold and is worth a sentence in
your write-up; `"review"` (functional form only) means the check produces a
plot for visual inspection rather than an automatic pass/fail, since
linearity-on-the-log-hazard-scale is a judgment call, not a threshold test.

## Diagnostic plots

```{r, fig.alt = "Proportional hazards test plot"}
plot(fit, which = "ph_plot")
```

```{r, fig.alt = "Influence diagnostics plot"}
plot(fit, which = "influence_plot")
```

```{r, fig.alt = "Functional form check for age"}
plot(fit, which = "functional_form_age")
```

## A real worked example: the NCCTG lung cancer trial

Simulated data proves the code runs; it doesn't prove the diagnostics catch
anything real. `survival::lung` -- built into the `survival` package, no
download required -- is a better test case: `ph.ecog` has a few missing
values that `coxph()` will otherwise drop silently, and `sex` is a
borderline case for the proportional hazards assumption in this dataset.

```{r}
library(survival)

lung2 <- lung
lung2$status <- lung2$status - 1  # recode 1/2 -> 0/1 event coding
lung2$sex <- factor(lung2$sex, labels = c("male", "female"))

# ph.ecog has a handful of NAs; drop them explicitly rather than letting
# coxph() do it silently, so the row count in the model is exactly what
# you'd report in a methods section.
lung2 <- lung2[stats::complete.cases(lung2[, c("time", "status", "sex", "age", "ph.ecog")]), ]

fit_lung <- fit_coxph_pipeline(
  data       = lung2,
  time_var   = "time",
  event_var  = "status",
  covariates = "sex",
  adjust_for = c("age", "ph.ecog")
)

fit_lung
```

```{r}
fit_lung$verdicts
```

```{r, fig.alt = "Proportional hazards test plot, lung data"}
plot(fit_lung, which = "ph_plot")
```

```{r, fig.alt = "Functional form check for ph.ecog, lung data"}
plot(fit_lung, which = "functional_form_ph.ecog")
```

Two things worth checking by eye rather than taking on faith: confirm
`fit_lung$model_adjusted$n` matches the row count of `lung2` after the
`complete.cases()` filter (so you know exactly which observations the
hazard ratios describe), and look at whether `sex` clears the proportional
hazards test here -- it's close enough in this dataset that it's worth a
sentence in a write-up either way, rather than a silent pass.

