---
title: "Getting Started: Epidemic Simulation Pipeline"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started: Epidemic Simulation Pipeline}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4
)
```

```{r setup}
library(omicsuite)
```

## Why stochastic, not deterministic

A deterministic SIR/SEIR model (a system of ODEs) gives you one trajectory:
given `beta` and `gamma`, the outbreak either happens or it doesn't, and if
it does, it always looks the same. Real outbreaks with a small number of
initial cases don't work that way -- by chance, the first infected
individual might recover before infecting anyone, and the whole thing fades
out even when R0 is comfortably above 1. `simulate_gillespie_epidemic()`
runs the *exact* stochastic process (Gillespie's direct method, no
time-discretization approximation) many times and shows you the spread of
outcomes, not just the average one.

## A basic SIR outbreak

```{r}
fit <- simulate_gillespie_epidemic(
  model = "SIR",
  initial_state = c(S = 999, I = 1, R = 0),
  params = list(beta = 0.4, gamma = 0.1),
  t_max = 100, n_sim = 200, seed = 1
)

fit
```

```{r, fig.alt = "SIR trajectory plot across realizations"}
plot(fit, which = "trajectory_plot")
```

Notice the grey lines: some of those 200 realizations barely got started
before recovering their way to extinction, even though `beta / gamma = 4`
is well above the epidemic threshold. That's the point of running an
ensemble instead of a single deterministic curve.

```{r, fig.alt = "Distribution of final outbreak size"}
plot(fit, which = "final_size_hist")
```

```{r}
fit$verdicts
```

## Near the epidemic threshold

This is where stochasticity matters most. With `R0` just above 1, whether
any given realization becomes a sustained outbreak or fizzles out is
genuinely uncertain -- a deterministic model would show exactly one
outcome, and it would hide that uncertainty entirely.

```{r}
fit_threshold <- simulate_gillespie_epidemic(
  model = "SIR",
  initial_state = c(S = 999, I = 1, R = 0),
  params = list(beta = 0.12, gamma = 0.1),
  t_max = 150, n_sim = 200, seed = 2
)

fit_threshold$r0
fit_threshold$prop_extinct
```

```{r, fig.alt = "Trajectory plot near the epidemic threshold"}
plot(fit_threshold, which = "trajectory_plot")
```

## Adding an incubation period: SEIR

If there's a meaningful delay between exposure and becoming infectious,
add an `E` compartment and a `sigma` rate (`1 / sigma` is the mean
incubation period):

```{r}
fit_seir <- simulate_gillespie_epidemic(
  model = "SEIR",
  initial_state = c(S = 999, E = 0, I = 1, R = 0),
  params = list(beta = 0.4, sigma = 0.2, gamma = 0.1),
  t_max = 150, n_sim = 200, seed = 3
)

fit_seir
```

```{r, fig.alt = "SEIR trajectory plot across realizations"}
plot(fit_seir, which = "trajectory_plot")
```

Compare `fit_seir$peak_time` to `fit$peak_time` from the SIR run above with
the same `beta` and `gamma` -- the incubation period delays the peak
without changing `R0`, since `R0 = beta / gamma` doesn't involve `sigma` at
all in this model structure. That's a useful sentence to have ready for a
methods section: the incubation period changes the outbreak's timing, not
its ultimate threshold behavior.

```{r, fig.alt = "Distribution of peak infection timing, SEIR"}
plot(fit_seir, which = "peak_time_hist")
```
