---
title: "Getting Started: Kaplan-Meier Pipeline"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started: Kaplan-Meier Pipeline}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4
)
```

```{r setup}
library(omicsuite)
```

## Kaplan-Meier vs. Cox: different questions

[`fit_coxph_pipeline()`](getting-started.html) models *how* covariates
change the hazard. `fit_km_pipeline()` answers a simpler, more descriptive
question first: what does survival actually look like in this sample, with
no modeling assumptions at all beyond independent censoring? In practice
you'd usually run this before, not instead of, the Cox pipeline -- it's the
plot that goes in a results section before the hazard ratios do.

## A simulated example

```{r}
set.seed(1)
n <- 200
dat <- data.frame(
  time  = rexp(n, rate = 0.05),
  event = rbinom(n, 1, 0.7),
  arm   = factor(sample(c("control", "treatment"), n, replace = TRUE))
)

fit <- fit_km_pipeline(dat, "time", "event", strata_var = "arm")
fit
```

## Reading the verdict table

```{r}
fit$verdicts
```

`interpretation[median_survival_<stratum>]` restates the median survival
time and its CI in plain language -- including the not-reached case
explicitly, rather than silently printing `NA`. `interpretation[log_rank_test]`
is the global test of whether the strata differ; `censoring_summary` is
worth checking before trusting either: heavy censoring is exactly when
"median not reached" starts showing up and when the log-rank test loses
power.

## The KM plot

```{r, fig.alt = "Kaplan-Meier survival curves by treatment arm"}
plot(fit, which = "km_plot")
```

If you have [`survminer`](https://cran.r-project.org/package=survminer)
installed, this includes a risk table underneath; without it, `omicsuite`
falls back to a plain `ggplot2` step-function plot so the pipeline still
works, just without the risk table and with a simpler (linearly
interpolated rather than stair-stepped) confidence ribbon.

## Optional: a parametric overlay

Sometimes you want to know whether a parametric family (Weibull,
exponential, ...) is a reasonable summary of the same data -- useful if
you're about to extrapolate beyond the observed follow-up period, which a
KM curve alone can't do.

```{r, fig.alt = "KM curve with a Weibull parametric overlay"}
fit_param <- fit_km_pipeline(dat, "time", "event", parametric_dist = "weibull")
plot(fit_param, which = "parametric_overlay")
```

```{r}
fit_param$verdicts[grepl("^parametric", fit_param$verdicts$check), ]
```

The AIC row is informational -- useful for comparing candidate
distributions against each other -- but the overlay plot is what actually
tells you whether the fit is any good; a low AIC among bad options is still
a bad fit, which is why that row is marked `"review"` rather than
auto-scored.
