---
title: "Getting Started: Multi-Omics Integration Pipeline"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started: Multi-Omics Integration Pipeline}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4
)
```

```{r setup}
library(omicsuite)
```

## Why not just PCA on the concatenated matrix?

Stacking a transcriptomics block with 20,000 features and a proteomics
block with 200 features into one matrix and running PCA gives you a result
dominated by whichever block has more features -- not necessarily the one
with more *signal*. `integrate_multiomics()` uses regularized generalized
canonical correlation analysis (RGCCA) instead, which finds components that
maximize covariation *between* blocks, so each layer contributes based on
how much it agrees with the others, not how large it is.

## A simulated example

```{r}
set.seed(1)
n <- 60
sample_ids <- paste0("patient_", seq_len(n))

# two omics layers with a shared latent signal on the first axis
transcriptomics <- matrix(
  rnorm(n * 30), nrow = n,
  dimnames = list(sample_ids, paste0("gene_", 1:30))
)
proteomics <- matrix(
  rnorm(n * 15), nrow = n,
  dimnames = list(sample_ids, paste0("protein_", 1:15))
)

subtype <- factor(rep(c("subtype_A", "subtype_B"), each = n / 2))
shared_signal <- ifelse(subtype == "subtype_A", 1.5, -1.5) + rnorm(n, sd = 0.5)
transcriptomics[, 1] <- transcriptomics[, 1] + shared_signal
proteomics[, 1] <- proteomics[, 1] + shared_signal

fit <- integrate_multiomics(
  blocks = list(transcriptomics = transcriptomics, proteomics = proteomics),
  group  = stats::setNames(as.character(subtype), sample_ids),
  ncomp  = 2,
  n_boot = 30,
  seed   = 1
)

fit
```

## Reading the verdict table

```{r}
fit$verdicts
```

`sample_alignment` confirms no samples were silently dropped when blocks
were matched by row name. `variance_explained[<block>]` is informational --
it tells you how much of each block's own variance the shared components
capture, which is a quick way to spot a block that isn't really
contributing to the integration. `loading_stability[<block>]` is the one
worth taking seriously before reporting a feature ranking: it's a
case-resampling bootstrap sign-agreement rate for each block's
top-loading features, not just a fit statistic.

## Do the shared components separate the subtypes?

```{r, fig.alt = "Sample scores on the first two shared components, colored by subtype"}
plot(fit, which = "block_scores")
```

```{r, fig.alt = "Variance explained per block"}
plot(fit, which = "variance_explained")
```

```{r, fig.alt = "Bootstrap loading stability by block"}
plot(fit, which = "stability")
```

```{r}
head(fit$stability[order(-fit$stability$stability), ])
```

The features at the top of that sorted table are the ones whose
contribution to the shared component survived resampling most
consistently -- a better place to start a follow-up biological
interpretation than the raw loading magnitude alone.
