---
title: "Getting Started: RNA-Seq Mixed Model Pipeline"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started: RNA-Seq Mixed Model Pipeline}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4,
  eval = FALSE  # model-fitting chunks are illustrative -- see note below
)
```

## Why not just fit each gene separately?

The traditional approach to RNA-seq differential expression -- a separate
negative binomial GLM per gene -- treats every gene as if it told you
nothing about any other gene. `fit_rnaseq_nb_pipeline()` instead fits one
multilevel model across all genes at once, with a gene-level random slope
for the effect of interest. Genes with little data (or noisy counts) get
pulled ("shrunk") toward the population-average effect; genes with strong,
consistent evidence keep their own estimate. This is the same partial-pooling
idea behind `lme4`/`brms` random effects, applied to differential expression.

**Note on this vignette:** the code chunks below are shown but not evaluated
during package build (`eval = FALSE`), because fitting even a small Stan
model takes longer than is reasonable for a vignette build step, and
`brms` is a `Suggests` dependency precisely so `omicsuite` doesn't force a
Stan toolchain on everyone who only wants the survival module. Copy these
chunks into an interactive session to run them for real.

```{r setup}
library(omicsuite)
library(brms)
```

## A simulated example

```{r}
set.seed(1)
n_genes <- 20
n_samples <- 16

dat <- expand.grid(
  gene   = paste0("gene_", seq_len(n_genes)),
  sample = paste0("sample_", seq_len(n_samples)),
  stringsAsFactors = FALSE
)
dat$condition <- factor(rep(
  rep(c("control", "treated"), each = n_samples / 2), n_genes
))
# a handful of genes get a real treatment effect; the rest are noise
responsive_genes <- paste0("gene_", 1:4)
dat$mu <- ifelse(
  dat$gene %in% responsive_genes & dat$condition == "treated",
  80, 40
)
dat$count <- stats::rnbinom(nrow(dat), mu = dat$mu, size = 4)

fit <- fit_rnaseq_nb_pipeline(
  data          = dat,
  gene_var      = "gene",
  count_var     = "count",
  condition_var = "condition",
  sample_var    = "sample",
  chains = 4, iter = 2000, warmup = 1000,
  cache_file = "rnaseq_vignette_fit.rds"
)

fit
```

## Reading the output

```{r}
fit$verdicts
```

`mcmc_convergence` is the one to check before trusting anything else --
if Rhat is elevated or the effective-sample-size ratio is low, more
iterations (or a reparameterized model) come before any biological
interpretation. `dispersion[shape]` is informational: a large shape
parameter means the negative binomial is behaving close to Poisson for
this dataset. `posterior_predictive_check` is a visual check, not an
automatic pass/fail -- see the plot below.

```{r, fig.alt = "Posterior predictive check plot"}
plot(fit, which = "pp_check")
```

```{r, fig.alt = "Rhat distribution across parameters"}
plot(fit, which = "rhat_plot")
```

## The shrinkage plot

This is the plot that makes the partial-pooling worth it: each gene's
condition effect, shrunk toward the population-level estimate (dashed
line) in proportion to how much evidence that gene provides.

```{r, fig.alt = "Gene-level shrinkage plot"}
plot(fit, which = "shrinkage_plot")
```

```{r}
head(fit$shrinkage[order(-abs(fit$shrinkage$gene_deviation)), ])
```

The genes furthest from the dashed line in the plot -- and at the top of
this sorted table -- are the ones with the strongest evidence of a
condition-specific effect after accounting for what every other gene in
the dataset is doing.
